---
title: "stringr 1.2.0"
date: "2017-02-18"
---

```{r setup, include = TRUE}
knitr::opts_chunk$set(comment = "#>", collapse = T)
library(stringr)
```

## API changes

This release includes a change to the API: `str_match_all()` now returns NA if an optional group doesn't match (previously it returned ""). This is more consistent with `str_match()` and other match failures.

```{r}
x <- c("a=1,b=2", "c=3", "d=")

x %>% str_match("(.)=(\\d)?")
x %>% str_match_all("(.)=(\\d)?,?")
```

## New features

There are three new features:

*   In `str_replace()`, `replacement` can now be a function that is called once
    for each match and who's return value is used to replace the match.
  
    ```{r}
    redact <- function(x) str_dup("-", str_length(x))
    
    x <- c("It cost $500", "We spent $1,200 on stickers")
    x %>% str_replace_all("\\$[0-9,]+", redact)
    ```

*   New `str_which()` mimics `grep()`:

    ```{r}
    fruit <- c("apple", "banana", "pear", "pinapple")
    
    # Matching positions    
    str_which(fruit, "p")

    # Matching values
    str_subset(fruit, "p")
    ```

*   A new vignette (`vignette("regular-expressions")`) describes the 
    details of the regular expressions supported by stringr.
    The main vignette (`vignette("stringr")`) has been updated to 
    give a high-level overview of the package.

## Minor improvements and bug fixes

There were three other minor improvements and bug fixes:

* `str_order()` and `str_sort()` gain explicit `numeric` argument for sorting
  mixed numbers and strings.

* `str_replace_all()` now throws an error if `replacement` is not a character
  vector. If `replacement` is `NA_character_` it replaces the complete string 
  with `NA`.

* All functions that take a locale (e.g. `str_to_lower()` and `str_sort()`)
  default to "en" (English) to ensure that the default is consistent across
  platforms.
